Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-06 analysis/paper/paper.Rmd
edits on intro and methods

<!--burial study- network analysis-->
Burials analysis has long developed in archaeology as an approach to study past social structure through mortuary practices, such as cultural and biological records in burial contexts, that can represent the social status or identity of the deceased [@Binford1971; @Saxe1970]. Despite the criticism for overlooking the burial manipulation by the living that does not necessarily connect to the real life [@Hodder1980; @Pearson1982], with comparisons between different archaeological contexts of death and life in mind, burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference for social structures [@Byrd1995; @Seikel2011]. To study the relations among burials, the recent development of network science provides many new tools to visualize and analyze relationships. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013]. That is, a network can be viewed as a patterned aggregation that includes individual elements, pair-wise relationships (the dyads), and overall structure showing global patterning represented in data.

A network is generally visualized as a graph consisting of a set of socially-relevant nodes, or actors, connected by edges representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, and co-presence of material culture based on theories for relationships [@Mills2017; @Brughmans2018]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarity, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial analysis. Burial goods, especially for high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status that can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization by identifying the relationships among burials.

Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanism. There are two common methodological approaches, one is directly examining network properties, such as node position and overall structures of networks built on archaeological data, to find meaningful patterns [@Mills2017]; another one is modeling theory to represent and test dependence of relationships [@Brughmans2018]. Node position is important to identify the role of a node in a network, such as its centrality, representing the individual influence or social prominence in a group, while network structure is helpful to generalize relationship patterns. Recent applications include long term inter-site relationship from the Epipalaeolithic to the early Neolithic in Near East according to trade items [@Coward2013], and hierarchical communication systems in the Kofun period in Japan through prestige goods [@Mizoguchi2013]. For modeling approaches, they provide a robust way to statistically test hypotheses about the relationships between actors to examine the structure of the network and major factors that affect the formation of the structure [@Freeman2004; @Salvini2010]. Examples include testing hunter-gatherer exchange network across the Kuril Islands based on ceramic compositions [@Erik2015], and modeling and comparing chronological and social hypotheses to explain the differences between burials [@Sosna2012]. This paper will use modeling approaches for the relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts.

<!--archaeology case-->
The archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and some accumulation patterns in residential area around the same time as European contact in the 17th century [@Chen2007], which might indicate the pursuit of prestige or wealth in those societies. The prestige goods include imported ornaments and porcelains that entered northeastern Taiwan via regional exchange networks involved the Europeans, Han Chinese traders, and Indigenous groups. The concept of prestige goods may develop earlier before the European arrival, but could be amplified during the European colonial period. For example, a person with more imported goods may be treated as higher status. A general model to summarize these case studies might be that the support of a colonial power combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals. In this study, we use a pericolonial model to explain the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century and the Chinese in the 19th century. Our purpose is to identify changes in network structure of burials before and after the foreign contacts, and test our hypothesis that increased inequality observed in social network after the European arrival. This is important to understand how an Indigenous society reacted to the presence of foreign imperial power.

We applied Bayesian SNA to burial data from Kiwulan, an Iron Age site in northeast Taiwan. We viewed individual burials as actors or nodes in the social network with ties drawn between burials when the presence of the same burial goods that can represent the same social group with similar economic or political status. We construct a network that represents higher social inequality and tested hypotheses that burial relations becomes more hierarchical after the European arrival. We ask the questions: 1. Do the observed burial data after the European presence shows more similar network quality as our hypothesized model? 2. Whether European colonial activities in 17 century Taiwan resulted in the emergence of social inequality in an Indigenous society can be detected by burials networks. 3. What is the major variables affecting or forming the higher degree of popularity (e.g. a few node has more relationship with other nodes) in the European contact period that might hint at social heterogeneity during this period. This study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups. A Bayesian approach enables the efficient quantification for uncertainty, parametrization, and model evaluation of social network metrics.

@@

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical models for networks that allows directly modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. ERGMs differs from traditional network approaches that edges are only formed according to their corresponding predictors separately without taken into account the interactions among edges [@Morris2008]. In ERGMs, networks are viewed as dependent variables, the formation of a tie is also influenced by network dependencies, and the attributes of nodes/edges [@Snijders2011]. For example, nodes with similar attributes is more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), or popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. We can model those network statistics as direct functions of ties, where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. By specifying the forms of configurations, we can build a hypothesis-based model to generate a distribution of random networks. Such distribution consists of a large number of possible networks that provides a statistical inference for an observed network [@Robins2007].  This helps to understand whether an observed network shows significantly more or less property of interest than the random networks we modeled.

ERGMs can describe a network by capturing the dependency structure by specifying parameters under assumptions and contexts, which is useful for exploring archaeological network. For instance, @Amati2019 demonstrated how to reconstruct of three different networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore their interaction mechanisms using ERGMs. Three network models test different hypotheses for the relationship between sites based on proximity, inter-cultural, and exchange network to explain the presence of hub sites they observed. They found that the presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of only one model exclusively. However, they also pointed out some limitations of ERGMs, such as static outcome and sensitivity to missing data. In addition, it is difficult in ERGMs to estimate model parameters and interpret the result due to the intractable normalizing constant and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is an important step to make probability distribution to integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. Degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. Those complex computational processes in ERGMs can be also resolved by implementing ERGMs in a Bayesian framework.

<!--Bayesian inference for Exponential Random Graph Models--> 
Bayesian approach to ERGMs is an efficient computational tool for social network analysis because it incorporates prior information about the network effects into the model and offers uncertainty quantification by evaluating the posterior distribution of the parameters associated with the network effects [@Caimo2015; @Nemmers2019]. This provides better estimations for complex social network models with heterogeneous data [@Caimo2017]. For archaeological data, missing data could be a problem leading to misinterpretation, which can be also reduced with Bayesian modeling that provides average 80% correct prediction for the ties and allow 25% false with a third of data missing [@Koskinen2010]. To overcome these limitations in ERGMs, Bayesian modeling applies Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm to estimate the double intractable posteriors [@Caimo2011]. MCMC algorithms can produce approximate maximum likelihood estimators and graph statistics for further specification of transitivity and heterogeneity, this enables ERGMs to deal with complicated dependence patterns with a ease [@Snijders2006]. This alleviate above computational problems and gives a better convergence results. By fitting an ERMG with MCMC algorithm, a Bayesian approach provides posterior probability that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability enables the estimation the effect of ERGM parameter by looking at the posterior mean and 95% credible intervals, which can replace p-value for accessing a null hypothesis [@Caimo2017bayesian]. The criteria is that the odds ratio greater than one means positive effect of ERGM parameter, while odds ratio less than one represents negative effect.

# Burial data from Kiwulan, NE Taiwan

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan site, an Iron Age settlement site (1350-1850 AD) located in northeastern Taiwan, which experienced the European colonial impacts in the early 17th century and the large wave of Chinese immigrants in the 19th century [@Chen2007]. The excavation revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with in-situ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison. Those burials are oriented in an east-west direction on the north side of the residential area indicated by post-holes and `in-situ` wooden posts, indicating a well organized space arrangement. Previous studies point out a presence of uneven distribution of prestige goods across burials without an agreement that whether this uneven distribution hints vertical social differences. @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian based on the comparative analysis of frequency and proportion of all burial goods. She found that the burials with high valued burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status.

The earliest record of the use of prestige goods in burials can be traced back to the time when the European arrival in the 17th century according to historical records from the Spanish [@Borao2009; @LiandWu2006]. For example, the document written by a Spanish Priest described that the Spanish soldier used carnelian beads to exchange natural resources with local Indigenous people since their high values in Indigenous culture [@LiandWu2006]. In addition, there was a custom that people would bury imported beads and ceramics with deceased to display their status and influence [@LiandWu2006]. Despite we observed the uneven distribution of prestige goods in burials, the degree of differentiation across burials over time remain unclear. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied. That is, the observed uneven distribution may be explained as a result of social change from less social inequality to more social inequality when Indigenous societies in Northeastern Taiwan were involved in the complex trade network stimulated by the Europeans. We assume that the changes would be supported by chronological differences in the structure of burial networks. Analyzing and comparing networks between different phases relative to cultural contact will give an insight into changes in social differences. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

<!--chronology-->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`) for network comparison based on established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon dates [@LWandBM2020ornament; @Chen2007]. We excluded the burials from the Chinese phase (n = `r chi`) for analysis due to the smaller sample size. There are eight burials heavily disturbed by modern construction that are excluded to avoid misinterpreting due to distorted chronology. We reconstructed social relations between individuals by linking them according to similar prestige goods. This is based on the assumption that the prestige goods imply social practices of trade, exchange, and gifting that reflect social relations between individuals [@Coward2013, pp. 252]. Also, prestige materials usually relate to a representation of social status because of their limited accessibility that can be used as status indicators [citations]. Therefore, we build networks where burials represent actors (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include agate beads, gold-foil beads, imported Chinese porcelains, gold foils, and fish-shaped ornaments that are considered as high-value items by studies across different archaeological contexts [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011] and the description in historical records [@Borao2009; @LiandWu2006].

@@

## data and hypothesis

We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. We assume that the pericolonial impact hypothesis would be supported by changes in network structures build upon flow of resources, prestige goods here. We treat the presence of relations between burials based on number or types of  prestige items from regional trade network. where a few prestige goods controlled by a small number of individuals. This is because prestige items could represent a relationship, where a few prestige goods controlled by a small number of individuals. Based on this assumption, we treat the presence of relations between burials through the burial goods in common.  Given that gold-foil beads and agate beads commonly found across burials with substantial differences in counts, we classified those beads into three levels, high, medium, and low, to indicate quantity based on the distribution of their counts. That means if burial 1 and burial 2 both have high quantity of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common. Node attributes here include osteological data and mortuary practices, such as age, gender, ritual practice, and overall burial goods quantity index. Ritual practice is identified by locally made ceramics placed above graves that suggests funeral feasting according to historical records. Quantity index is an important attribute for economical inequality that we assigned the burials into three classes, high, medium, and low, as an index of wealth, based on the total number of burial goods. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

>>>>

<!--burial study- network analysis-->
Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviours can represent the social status or identity of the deceased [@Binford1971; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past social organisation [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. The recent development of network science provides many new tools to visualize and analyze relationships among ancient burials. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

<!-- intro to networks -->
A network is generally visualized as a graph consisting of a set of socially-relevant nodes, or actors, connected by edges representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, and co-presence of material culture based on theories of relationships [@Mills2017; @Brughmans2018]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarity, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial contexts. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

<!-- two types of network analysis -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches, one is directly examining network properties, such as node position and overall structures of networks built on archaeological data, to find meaningful patterns [@Mills2017]. Another approach is using models and simulations to represent and test dependence of relationships [@Brughmans2018]. Node position is important to identify the role of a node in a network, such as its centrality, representing the individual influence or social prominence in a group, while network structure is helpful to generalize relationship patterns. Recent applications include the investigation of long term inter-site relationships from the Epipalaeolithic to the early Neolithic in the Near East according to trade items [@Coward2013]. Similarly, @Mizoguchi2013 analysed hierarchical communication systems in the Kofun period in Japan through the study of prestige goods. For the second approach, researchers typically statistically test hypotheses about the relationships between actors to examine the structure of the network and infer major factors that affect the formation of the structure [@Freeman2004; @Salvini2010]. Examples include testing hunter-gatherer exchange networks across the Kuril Islands based on ceramic composition [@Erik2015].  <!-- better to have a specific case study by a scholar who is likely to peer review this paper, try to find some recent publication --> . This paper will use a novel Bayesian modeling approach to investigate the relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organisation.

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and some accumulation patterns in residential areas around the same time as European contact in the 17th century [@Chen2007]. This might indicate the pursuit of prestige or wealth in those Indigenous societies. Prestige goods include imported ornaments and porcelains that entered northeastern Taiwan via regional exchange networks that involved the Dutch, Spanish, Han Chinese traders, and local Indigenous groups. The use of prestige goods may have began in local Indigenous communities before a European presence became established, but was amplified during the European colonial period <!-- citation -->. For example, an Indigenous person possessing more imported goods may have been recognised as having higher status by their community <!-- citation to some historical record -->. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals, and unequal access to trade networks <!-- citation -->.

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century <!-- citation -->. Our purpose is to examine the network structure of burials before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the of network our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups.

@@

<!-- current approaches to network analysis: important to show how what we are doing is amazing and new and not done before by other archaeologists. We need a sentence or two to say what do archaeologists usually do for network statistics... and make it clear that we are doing something new -->
Exponential Random Graph Models (ERGMs) are an important family of statistical models for networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. ERGMs differs from traditional network approaches in that edges are only formed according to their corresponding predictors separately without taking into account the interactions among edges [@Morris2008]. In ERGMs, networks are viewed as dependent variables, the formation of a tie is also influenced by network dependencies, and the attributes of nodes/edges [@Snijders2011]. For example, nodes with similar attributes are more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration, that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), or popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. We can model those network statistics as direct functions of ties, where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. By specifying the forms of configurations, we can build a hypothesis-based model to generate a distribution of random networks that represent our model. Such distribution consists of a large number of possible networks that enables statistical inference and comparison with an observed network [@Robins2007]. This helps us understand whether an observed network shows significantly more or less of a property of interest than the random networks generated from our model assumptions.

<!-- ERGM in archaeology, and limitations --> 
To date, ERGM has only been used once in archaeology, by @Amati2019 who reconstructed three different networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore their interaction mechanisms. These network models evaluated different hypotheses about the importance of three variables, proximity, inter-cultural items, and pottery types, on the presence of hub sites they observed. They found that the presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. They also pointed out some limitations of ERGMs, such as static outcomes and sensitivity to missing data. In addition, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is an important step to make probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. Degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are efficient computational tools for social network analysis because they incorporates prior information about the network configurations into the model and offers uncertainty quantification by evaluating the posterior distributions of the parameters associated with network configurations [@Caimo2015; @Nemmers2019]. This provides better estimations for complex social network models with heterogeneous data [@Caimo2017]. For archaeological studies, missing data is often a problem leading to misinterpretation. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling to provides an average of 80% correct predictions for the ties, and allow 25% false predictions with a third of data missing [@Koskinen2010] <!-- maybe a little more detail and clarification needed here -->. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm to estimate the double intractable posteriors [@Caimo2011]. MCMC algorithms can produce approximate maximum likelihood estimators and network statistics for further specification of transitivity and heterogeneity. The use of MCMC enables ERGMs to deal with complicated dependence patterns with ease [@Snijders2006]. This alleviates the above computational problems and gives a better convergence results. By fitting an ERMG with an MCMC algorithm, a Bayesian approach provides posterior probabilities that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability enables the estimation the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect.

# Burial data from Kiwulan, Northeastern Taiwan

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) located in northeastern Taiwan, which experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison <!-- fig reference to a map of burials at the site-->. Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organiation with post-European social organisation. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period.

<!-- take the citations from this paragraph and add them to one of the intro paras, them delete this para -->
The earliest record of the use of prestige goods in burials can be traced back to the time when the European arrival in the 17th century according to historical records from the Spanish [@Borao2009; @LiandWu2006]. For example, the document written by a Spanish Priest described that the Spanish soldier used carnelian beads to exchange natural resources with local Indigenous people since their high values in Indigenous culture [@LiandWu2006]. In addition, there was a custom that people would bury imported beads and ceramics with deceased to display their status and influence [@LiandWu2006]. Despite we observed the uneven distribution of prestige goods in burials, the degree of differentiation across burials over time remain unclear. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied. That is, the observed uneven distribution may be explained as a result of social change from less social inequality to more social inequality when Indigenous societies in Northeastern Taiwan were involved in the complex trade network stimulated by the Europeans. We assume that the changes would be supported by chronological differences in the structure of burial networks. Analyzing and comparing networks between different phases relative to cultural contact will give an insight into changes in social differences. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

<!--chronology-->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [@LWandBM2020ornament; @Chen2007]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. There are eight burials heavily disturbed by modern construction that are also excluded because we cannot determine their chronology. We reconstructed social relations between individuals by linking them according to similar prestige goods [cf. @Coward2013, pp. 252].Therefore, we built networks where burials represent actors (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include agate beads, gold-foil beads, imported Chinese porcelains, gold foils, and fish-shaped ornaments. Previous studies indicate that these items are considered as high-value across different archaeological contexts [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. Historical accounts also support these items as prestige goods [@Borao2009; @LiandWu2006].

@@

## Data and Hypothesis

<!-- we should tell the reader what does it mean high-medium-low --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Gold-foil beads and agate beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of three levels, high (50-1000?), medium (20-50?), and low (1-10?), of these beads. That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age, sex, and ritual pottery, and an overall burial goods values... index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The quantity index is an important attribute for economical inequality that we assigned the burials into three classes, high, medium, and low, as an index of wealth, based on the total number of burial goods. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

<!-- stopped editing here -->
